---
title: "README"
output: 
  html_document:
    keep_md: yes
    code_folding: hide
    toc: true
    toc_float: true
author: "Marcus W. Beck, beck.marcus@epa.gov, Janne Alahuhta, Janne.Alahuhta@oulu.fi"
---

```{r warning = F, cache = F, message = F}
library(ggplot2)
library(maptools)
library(gridExtra)
library(RColorBrewer)
library(dplyr) 
library(gridExtra)
library(mgcv)
source('R/funcs.R')
```

### Files

All files available [here](https://github.com/fawda123/veggam/tree/master/data)

* `ecoregs.RData` SpatialPolygonsDataFrame for MN ecoregions
* `mncounties.RData` SpatialPolygonsDataFrame for MN counties
* `mnstate.RData` SpatialPolygonsDataFrame for state outline
* `veg_dat.RData` DNR vegetation transect data from 1992 to 2014, created in `fishveg.RProj`, combination of nrri processed data, current transect data, and gaps filled with another file
* `veg_rch.RData` Summarized vegetation richness data from veg_dat, includes total richness (`richtot`), submersed species richness (`richsub`), presence/absence of different species (as 1/0), richness of native submersed species (`richnat`), covariates, and categorical describing occurrence of p. crispus and m. spicatum (none, either, or both). Scientific names were manually verified during processing to remove duplicates (all species had scientific names except 'filamentous algae', 'planktonic algae', and 'Fern group', these were removed - 2642 records out of 894920 ~0.3% of total)

### Species maps

Maps showing species richness of submersed macrophytes from MNDNR transect database (top) and presence/absence of the invasives M. spicatum and P. crispus (bottom).
```{r fig.height = 7, fig.width = 7, echo = F, fig.cap = 'Species richness of submersed macrophytes from MNDNR transect database.', message = F, cache = T, warning = F}

# raw veg, spatial data
data(veg_rch)
data(ecoregs)
data(mncounties)
data(mnstate)

# prep spatial data
mnstate <- fortify(mnstate)
ecoregs <- fortify(ecoregs)
mncounties <- fortify(mncounties)
ecoregs$Ecoregion <- factor(ecoregs$id, levels = c(6, 2, 5, 0, 4, 3, 1), # ided from arcmap
  labels = c('DA', 'LAP', 'NCHF', 'NGP', 'NLF', 'NMW', 'WCBP')
)

# subm rich as factor
toplo <- veg_rch %>% 
   mutate(richsub = cut(richsub, breaks = c(-Inf, 5, 10, 15, 20, Inf), 
    labels = c('<5', '<10', '<15', '<20', '>20'))
    ) %>% 
  data.frame
labs <- list(expression(phantom('')<=5), expression(phantom('')<=10), expression(phantom('')<=15), expression(phantom('')<=20), expression(phantom('')>20))

# base map for plots
pbase <- ggplot(mncounties, aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group), fill = NA, colour = 'grey') +
  geom_polygon(data = mnstate, aes(x = long, y = lat), fill = NA, colour = 'black') + 
  geom_polygon(data = ecoregs, aes(x = long, y = lat, group = group, fill= Ecoregion),  
    alpha = 0.5) +
  theme_classic() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
    axis.text.y=element_blank(),axis.ticks=element_blank(),
    axis.title.x=element_blank(),
    axis.title.y=element_blank(),
    panel.background=element_blank(),panel.border=element_blank(),
    panel.grid.major=element_blank(),
    panel.grid.minor=element_blank(),plot.background=element_blank(), 
    legend.box.just = 'right'
    ) +
  coord_equal()

# add richness data to plot
pbase + 
  geom_point(data = toplo, aes(x = utmx, y = utmy, 
    size = richsub), fill = 'white', colour = 'black', alpha = 0.8, shape = 21, stroke = 1) +
  scale_size_discrete('Richness', range = c(1, 8), labels = labs) +
  scale_fill_manual(values = brewer.pal(9, 'Spectral')[c(1, 2, 3, 5, 7, 8, 9)])
```
```{r fig.cap = 'Distribuion of lakes with P. crispus, M. spicatum, both, or none.', message = F, cache = T, warning = F, message = F,fig.height = 7, fig.width = 7, echo = F}
# p/a of invasives as facet map
pbase + 
  geom_point(data = toplo, aes(x = utmx, y = utmy, shape = Invasive), fill = 'white', size = 3, alpha = 0.8) +
  scale_shape_manual(values = c(21, 22, 23, 24)) + 
  facet_wrap(~Invasive) +
  scale_fill_manual(values = brewer.pal(9, 'Spectral')[c(1, 2, 3, 5, 7, 8, 9)])

```

### Example GAM: submersed richess

The first model shows a tensor product smooth fit for submersed species richness as a function of lake depth and total phosphorus, with separate smooths for presence/absence of invasives.  The lakes are categorized as no invasives, ms (M. spicatum), pc (P. crispus), or both.  Submersed species richness is modelled as a Poisson-distributed variable. Each 3d plot shows the modelled response of richness as a function of depth and phosphorus for each of the four levels of invasive categories (none, ms, pc, both). 
```{r eval = T, message = F}
data(veg_rch)

tomod <- select(veg_rch, dow, date, Invasive, richsub, pa, depthm, totalp) %>% 
  group_by(dow) %>% 
  filter(date == max(date)) %>% 
  ungroup %>% 
  na.omit

richmod <- gam(richsub ~ 
  te(depthm, totalp, by = Invasive, bs = c("ts")), 
  family = poisson(link = 'log'),
  data = tomod
  )

theta <- 45; phi <- 25; n.grid <- 50; color <- 'topo'; type <- 'link'; zlab <- 'Subm. richness'
vis.gam(richmod, theta = theta, phi = phi, n.grid = n.grid, color = color, type = type, cond = list(Invasive = 'none'), zlab = zlab)
vis.gam(richmod, theta = theta, phi = phi, n.grid = n.grid, color = color, type = type, cond = list(Invasive = 'ms'), zlab = zlab)
vis.gam(richmod, theta = theta, phi = phi, n.grid = n.grid, color = color, type = type, cond = list(Invasive = 'pc'), zlab = zlab)
vis.gam(richmod, theta = theta, phi = phi, n.grid = n.grid, color = color, type = type, cond = list(Invasive = 'both'), zlab = zlab)
```

These plots show an alternative view of the results for each explanatory variable.
```{r, fig.height = 8, fig.width = 10}

# get model predictions at different levels of input data
evals <- c('totalp', 'depthm')
consts <- c('depthm', 'totalp')
levs <- c('none', 'ms', 'pc', 'both')
qconsts <- c(0.25, 0.75)
grds <- expand.grid(evals, consts, levs, qconsts, stringsAsFactors = F)
grds <- grds[grds[, 1] != grds[, 2], ]
grds$i <- 1:nrow(grds)
names(grds) <- c('evals', 'consts', 'levs', 'qconsts', 'i')

resvals <- vector('list', length = nrow(grds))
for(i in 1:nrow(grds)){
  
  evali <- grds[i, 'evals']
  consti <- grds[i, 'consts']
  levsi <- grds[i, 'levs']
  qconsti <- grds[i, 'qconsts']
  
  res <- pred_gam(
    richmod, 
    tomod, 
    eval = evali, 
    const = consti, 
    fac = 'Invasive',
    lev = levsi, 
    qconst = qconsti, 
    n = 100, 
    se.fit = T, 
    type = 'response'
    )
  resvals[[i]] <- res
  
}
resvals <- do.call('rbind', resvals)

toplo1 <- filter(resvals, eval == 'totalp')
toplo2 <- filter(resvals, eval == 'depthm')
p1 <- ggplot(toplo1, aes(x = var, y = fit)) + 
  geom_line() +
  geom_line(aes(y = fit + se.fit), linetype = 'dashed') +
  geom_line(aes(y = fit - se.fit), linetype = 'dashed') + 
  scale_y_continuous('Submersed species richness', limits = c(-5, 35)) +
  facet_grid(Invasive ~ qconst) + 
  theme_bw() + 
  scale_x_continuous('Mean total phosphorus') + 
  ggtitle('(a) Constant depth')

p2 <- ggplot(toplo2, aes(x = var, y = fit)) + 
  geom_line() +
  geom_line(aes(y = fit + se.fit), linetype = 'dashed') +
  geom_line(aes(y = fit - se.fit), linetype = 'dashed') + 
  scale_y_continuous('Submersed species richness', limits = c(-5, 35)) +
  facet_grid(Invasive ~ qconst) + 
  theme_bw() + 
  scale_x_continuous('Maximum lake depth') +
  ggtitle('(b) Constant phosphorus')

grid.arrange(p1, p2, ncol = 2)
```

### Example GAM: p/a largeleaf pondweed

The second model is the same as the first except p/a of largeleaf pondweed is modelled as a binomial (logit link) variable.  The result shows the conditional probability of presence for largelead pondweed given depth, phosphorus, and p/a of invasives. 
```{r}

data(veg_rch)

tomod <- select(veg_rch, dow, date, Invasive, richsub, pa, depthm, totalp) %>% 
  group_by(dow) %>% 
  filter(date == max(date)) %>% 
  ungroup %>% 
  na.omit

pamod <- gam(pa ~ 
  te(depthm, totalp, by = Invasive, bs = c("ts")), 
  family = binomial(link = 'logit'),
  data = tomod
  )

theta <- 45; phi <- 25; n.grid <- 50; color <- 'topo'; type <- 'link'; zlab <- 'Prob. of P. amplifolius'
vis.gam(pamod, theta = theta, phi = phi, n.grid = n.grid, color = color, type = type, cond = list(Invasive = 'none'), zlab = zlab)
vis.gam(pamod, theta = theta, phi = phi, n.grid = n.grid, color = color, type = type, cond = list(Invasive = 'ms'), zlab = zlab)
vis.gam(pamod, theta = theta, phi = phi, n.grid = n.grid, color = color, type = type, cond = list(Invasive = 'pc'), zlab = zlab)
vis.gam(pamod, theta = theta, phi = phi, n.grid = n.grid, color = color, type = type, cond = list(Invasive = 'both'), zlab = zlab)
```
These plots show an alternative view of the results for each explanatory variable.
```{r, fig.height = 8, fig.width = 10, warning = F}

# get model predictions at different levels of input data
evals <- c('totalp', 'depthm')
consts <- c('depthm', 'totalp')
levs <- c('none', 'ms', 'pc', 'both')
qconsts <- c(0.25, 0.75)
grds <- expand.grid(evals, consts, levs, qconsts, stringsAsFactors = F)
grds <- grds[grds[, 1] != grds[, 2], ]
grds$i <- 1:nrow(grds)
names(grds) <- c('evals', 'consts', 'levs', 'qconsts', 'i')

resvals <- vector('list', length = nrow(grds))
for(i in 1:nrow(grds)){
  
  evali <- grds[i, 'evals']
  consti <- grds[i, 'consts']
  levsi <- grds[i, 'levs']
  qconsti <- grds[i, 'qconsts']
  
  res <- pred_gam(
    pamod, 
    tomod, 
    eval = evali, 
    const = consti, 
    fac = 'Invasive',
    lev = levsi, 
    qconst = qconsti, 
    n = 100, 
    se.fit = T, 
    type = 'response'
    )
  resvals[[i]] <- res
  
}
resvals <- do.call('rbind', resvals)

toplo1 <- filter(resvals, eval == 'totalp')
toplo2 <- filter(resvals, eval == 'depthm')
p1 <- ggplot(toplo1, aes(x = var, y = fit)) + 
  geom_line() +
  geom_line(aes(y = fit + se.fit), linetype = 'dashed') +
  geom_line(aes(y = fit - se.fit), linetype = 'dashed') + 
  scale_y_continuous('Probabiliyt of P. amplifolius', limits = c(0, 1)) +
  facet_grid(Invasive ~ qconst) + 
  theme_bw() + 
  scale_x_continuous('Mean total phosphorus')+ 
  ggtitle('(a) Constant depth')

p2 <- ggplot(toplo2, aes(x = var, y = fit)) + 
  geom_line() +
  geom_line(aes(y = fit + se.fit), linetype = 'dashed') +
  geom_line(aes(y = fit - se.fit), linetype = 'dashed') + 
  scale_y_continuous('Probability of P. amplifolius', limits = c(0, 1)) +
  facet_grid(Invasive ~ qconst) + 
  theme_bw() + 
  scale_x_continuous('Maximum lake depth') +
  ggtitle('(b) Constant phosphorus')

grid.arrange(p1, p2, ncol = 2)

```